---
title: "Wine Quality Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Wine Quality Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

# Understanding the Dataset (Exploratory Data Analysis (EDA))

## Loading the Dataset

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/yasserh/wine-quality-dataset\>*

### Reference:

*\<Kaggle. (n.d.). Wine Quality Dataset. Retrieved from https://www.kaggle.com/datasets/yasserh/wine-quality-dataset\>\

Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Exploratory Data Analysis
## Load dataset
```{r load dataset}
# Load the wine quality dataset
wine_data <- read.csv("WineQT.csv", colClasses = c(
  fixed_acidity = "numeric",
  volatile_acidity = "numeric",
  citric_acid = "numeric",
  residual_sugar = "numeric",
  chlorides = "numeric",
  free_sulfur_dioxide = "numeric",
  total_sulfur_dioxide = "numeric",
  density = "numeric",
  pH = "numeric",
  sulphates = "numeric",
  alcohol = "numeric",
  quality = "numeric",
  Id = "character"
), header = TRUE)

# Display the structure of the dataset
str(wine_data)

# View the first few rows of the dataset
head(wine_data)

# Remove the last column
wine_data <- wine_data[, -ncol(wine_data)]

# Open the dataset in a viewer window
View(wine_data)

```

## Measures of frequency
```{r}
# Calculate measures of frequency
summary_stats <- summary(wine_data)

# Display the measures of frequency
print(summary_stats)
```

## Measures of Central Tendency
```{r}
# Calculate measures of central tendency
means <- colMeans(wine_data, na.rm = TRUE)
medians <- apply(wine_data, 2, median, na.rm = TRUE)

# Calculate mode
getmode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}
modes <- sapply(wine_data, getmode)

# Create a data frame to display the measures
central_tendency <- data.frame(
  Variable = names(wine_data),
  Mean = means,
  Median = medians,
  Mode = modes
)

# Print the measures of central tendency
print(central_tendency)
```

## Measures of Distribution
```{r}
# Calculate measures of distribution
ranges <- apply(wine_data, 2, function(x) diff(range(x, na.rm = TRUE)))
variances <- apply(wine_data, 2, var, na.rm = TRUE)
standard_deviations <- apply(wine_data, 2, sd, na.rm = TRUE)

# Create a data frame to display the measures
distribution_measures <- data.frame(
  Variable = names(wine_data),
  Range = ranges,
  Variance = variances,
  Standard_Deviation = standard_deviations
)

# Print the measures of distribution
print(distribution_measures)
```

## Measures of Relationship
```{r}
# Calculate measures of relationship (correlation)
correlation_matrix <- cor(wine_data, use = "pairwise.complete.obs")

# Print the correlation matrix
print(correlation_matrix)
```

## Plots
```{r}
#Load package
library(ggplot2)

# Plot for 'fixed acidity'
ggplot(wine_data, aes(x = fixed_acidity)) +
  geom_histogram(binwidth = 0.5, fill = "skyblue", color = "black") +
  labs(title = "Histogram of Fixed Acidity", x = "Fixed Acidity", y = "Frequency")

# Plot for 'volatile acidity'
ggplot(wine_data, aes(x = volatile_acidity)) +
  geom_histogram(binwidth = 0.05, fill = "lightgreen", color = "black") +
  labs(title = "Histogram of Volatile Acidity", x = "Volatile Acidity", y = "Frequency")

# Plot for 'citric acid'
ggplot(wine_data, aes(x = citric_acid)) +
  geom_histogram(binwidth = 0.05, fill = "lightcoral", color = "black") +
  labs(title = "Histogram of Citric Acid", x = "Citric Acid", y = "Frequency")

# Plot for 'residual sugar'
ggplot(wine_data, aes(x = residual_sugar)) +
  geom_histogram(binwidth = 1, fill = "lightyellow", color = "black") +
  labs(title = "Histogram of Residual Sugar", x = "Residual Sugar", y = "Frequency")

# Scatter plot for 'fixed acidity' vs 'volatile acidity'
ggplot(wine_data, aes(x = fixed_acidity, y = volatile_acidity)) +
  geom_point(color = "blue") +
  labs(title = "Scatter Plot: Fixed Acidity vs Volatile Acidity", x = "Fixed Acidity", y = "Volatile Acidity")

# Scatter plot for 'fixed acidity' vs 'citric acid'
ggplot(wine_data, aes(x = fixed_acidity, y = citric_acid)) +
  geom_point(color = "green") +
  labs(title = "Scatter Plot: Fixed Acidity vs Citric Acid", x = "Fixed Acidity", y = "Citric Acid")

# Scatter plot for 'fixed acidity' vs 'residual sugar'
ggplot(wine_data, aes(x = fixed_acidity, y = residual_sugar)) +
  geom_point(color = "red") +
  labs(title = "Scatter Plot: Fixed Acidity vs Residual Sugar", x = "Fixed Acidity", y = "Residual Sugar")

```

# Preprocessing and Data Transformation
## Missing values
```{r}
# Check for missing values
missing_values <- anyNA(wine_data)

# Display result
if (missing_values) {
  print("The dataset contains missing values.")
} else {
  print("The dataset does not contain missing values.")
}
```

## Transformation
```{r}
# Select only numeric variables for transformation
numeric_vars <- wine_data[, sapply(wine_data, is.numeric)]

# Perform standardization (z-score normalization) on numeric variables
standardized_data <- scale(numeric_vars)

# Convert the standardized data back to a data frame
standardized_df <- as.data.frame(standardized_data)

# Replace the numeric variables in the original dataset with the standardized ones
wine_data[, sapply(wine_data, is.numeric)] <- standardized_df

# View the transformed dataset
head(wine_data)
```

# Training Model
## Data splitting
```{r}
# Load the required library
library(caret)

# Set seed for reproducibility
set.seed(123)

# Split the dataset into training and testing sets (70% training, 30% testing)
train_indices <- createDataPartition(wine_data$quality, p = 0.7, list = FALSE)
train_data <- wine_data[train_indices, ]
test_data <- wine_data[-train_indices, ]

# Summary of the split
cat("Training set size:", nrow(train_data), "\n")
cat("Number of variables in training set:", ncol(train_data), "\n")
cat("\n")
cat("Testing set size:", nrow(test_data), "\n")
cat("Number of variables in testing set:", ncol(test_data), "\n")
```

## Bootstrapping
```{r}
# Set seed for reproducibility
set.seed(123)

# Number of bootstrap iterations
num_iterations <- 1000

# Function to calculate the statistic of interest (e.g., mean, median) from a sample
calculate_statistic <- function(data) {
  # Example: Calculate the mean of the 'quality' variable
  return(mean(data$quality))
}

# Perform bootstrapping
bootstrap_results <- replicate(num_iterations, {
  # Sample with replacement
  boot_sample <- wine_data[sample(nrow(wine_data), replace = TRUE), ]
  # Calculate the statistic of interest
  calculate_statistic(boot_sample)
})

# Summary of bootstrapped results
summary(bootstrap_results)
```

## Cross-validation
```{r}
# Load the required library
library(caret)

# Define the number of folds
num_folds <- 10  # You can adjust this value based on your preference

# Define the control parameters for cross-validation
train_control <- trainControl(method = "cv", number = num_folds)

# Define the model training
# Example: Linear regression using the 'lm' function
model <- train(quality ~ ., data = wine_data, method = "lm", trControl = train_control)

# Display the model performance
print(model)
```

## Training different models
```{r}
# Load the required libraries
library(caret)
library(randomForest)

# Set seed for reproducibility
set.seed(123)

# Define the control parameters for cross-validation
train_control <- trainControl(method = "cv", number = 10)

# Model training - Linear Regression
linear_model <- train(quality ~ ., data = wine_data, method = "lm", trControl = train_control)

# Model training - Random Forest
rf_model <- train(quality ~ ., data = wine_data, method = "rf", trControl = train_control)

# Display model performance
print(linear_model)
print(rf_model)
```

## Model performance comparison
```{r}
# Compare model performance using resamples
model_comparison <- resamples(list(LinearRegression = linear_model, RandomForest = rf_model))

# Summarize the results
summary(model_comparison)
```

## Saving Model
```{r}
# Saving the best Random Forest model
saveRDS(rf_model$finalModel, "./models/rf_model.rds")

# Load the saved model
loaded_rf_model <- readRDS("./models/rf_model.rds")

# Prepare new data for prediction
new_data <- data.frame(
  fixed_acidity = 7.4,
  volatile_acidity = 0.7,
  citric_acid = 0.0,
  residual_sugar = 1.9,
  chlorides = 0.076,
  free_sulfur_dioxide = 11,
  total_sulfur_dioxide = 34,
  density = 0.9978,
  pH = 3.51,
  sulphates = 0.56,
  alcohol = 9.4
)

# Use the loaded model to make predictions
predictions_loaded_model <- predict(loaded_rf_model, newdata = new_data)

# Print predictions
print(predictions_loaded_model)

```


